Is essentiality of genes conserved in Enterobacteriaceae?

### 2017-06-20 documents/method-comparison.tex
Added an ambiguous region between non-essential and beneficial loss, used eggNOG to study beneficial losses, generated a table containing all KEGG pathways with core essential and ancestrally essential genes

% adjust caption style
\usepackage[font=small,aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

@@

% this is required to include graphics
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{pdflscape}
\graphicspath{{../figures/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar
\usepackage{amsmath}

% document begins here
\begin{document}
\vspace*{0.35in}

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Title}
}
\newline
% authors go here:
\\
%Fatemeh Ashari Ghomi\textsuperscript{1},
%Paul P. Gardner\textsuperscript{1,*},
%Lars Barquist\textsuperscript{2}
%\\
%\bigskip
%\bf{1} University of Canterbury
%\\
%\bf{2} Wurzburg University
%\\
%\bigskip
%* paul.gardner@canterbury.ac.nz

\end{flushleft}

\section*{Abstract}
%Many genes have been identified with advances in sequencing technology and genome annotation methods. However, not all of these genes are of the same importance. We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae. We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes. After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level. Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.

% now start line numbers
%\linenumbers

% the * after section prevents numbering
\section{Introduction}
%The study of cells can give us new insights for understanding the foundation of life. However, the naturally existing cells are very complex which hampers the study of their constituents. Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.

Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions. Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes. Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}. Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}. Infectious diseases are among the top major causes of mortality worldwide. Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}. Therefore, it is urgent to find new drugs for infectious diseases. New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease. Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]

In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}. With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}. Therefore, the use of experimental methods for the identification of essential genes is vital. Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods. Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12. The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes. Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}. In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential. Both of these methods are labour intensive and are dependent on the accuracy of genome annotations. Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}. These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}. Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis. After a growth phase, mutants that are fitter outnumber the less fit ones. Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not. A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.

There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms. Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}. On the other hand, accessory essential genes are helpful in the study of specific lineages. Accessory essential genes may be useful in species-specific antibiotic discovery. Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.

%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
Freed et al. \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}. These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and

%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes. Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}. For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential. Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}. These include genes involved in replication and genes related to ribosome and its accessory proteins. The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S. typhi} and 10 genes vice versa. Barquist et al. \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113. 
%THIS IS IMPORTANT:
These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis. Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors. Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.

Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes. We are going to extend these studies to a larger scale in Enterobacteriaceae family. Our aim is to study the essentiality of genes in in 16 different organisms. These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.

%WHAT FATEMEH DID....

%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

%\begin{figure}
%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
%\caption{The species tree containing the 16 genomes used in this study. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
%\label{fig:species-tree}
%\end{figure}

\section{Results and Discussion}
\subsection{Incorporating multiple measures of gene essentiality improves prediction}
In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
\subsubsection{Comparison of old methods}
A number of methods have been used for evaluating the essentiality of genes using transposon insertion data. Freed et al. \cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes. These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length. Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.

Barquist et al. \cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes. Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential). They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.

Turner et al. \cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.

We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling. To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}. The results are depicted in Figure~\ref{fig:benchmark}. Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods. Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
\subsubsection{PCA based method}\label{sec:essentiality-call}
We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region. These predictors were selected as they were more accurate and less dependent on each other. Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component. As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods. We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}. Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation. We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper. The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential. The rest are non-essential genes. The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95. Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=1, scale=0.4]{essential-call-comparison-BW25113.pdf}
  \caption{}
  \label{fig:benchmark}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=10, scale=0.4]{essential-call-comparison-BW25113.pdf}
  \caption{}
  \label{fig:pca}
\end{subfigure}
\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is. (b) The distribution for our proposed method (NPEQ)}
\label{fig:essentiality-call}
\end{figure}

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
%\end{tabular}
%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
%\label{fig:benchmark}
%\end{figure*}

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BN373.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BW25113.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-CS17.pdf}\\
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-EC958.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ENC.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ERS227112.pdf}\\
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ETEC.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-NCTC13441.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ROD.pdf}\\
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SEN.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL1344.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL3261.pdf}\\
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STM.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STMMW.pdf}&
%\includegraphics[page=10, scale=0.22]{essential-call-comparison-t.pdf}\\
%\end{tabular}
%\caption{The plots show the distribution of our new prediction method which is calculated by running PCA on insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region and selecting first principal component. This method is called NPEQ.}
%\label{fig:pca}
%\end{figure*}
\subsection{TraDIS data is biased}
In Section~\ref{sec:essentiality-call} we proposed NPEQ method that quantifies the level of essentiality of a gene. However, if the transposon insertion is biased to specific regions in the genome, it can increase/decrease the values predicted by our method for genes and put them in a different level of essentiality. Different articles have reported biases in transposon insertion using Tn5 \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, canals_high-throughput_2012, langridge_simultaneous_2009}. We performed a detailed study of these biases. The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
\subsubsection{Distance from the origin of replication bias}
While a study has reported Tn5 insertion bias near the origin of replication \cite{barquist_comparison_2013}, another study has reported no bias \cite{rubin_essential_2015}. To study the bias towards the position of gene within genome, we plotted NPEQ for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}. The figure indicates that NPEQ increases when the genes are located further from the origin of replication. When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away due to the initiation of different replication forks. This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
%One might argue that the bias that we have observed may have happened due to the abundance of essential genes near the origin of replication. Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[page=6, scale=0.25]{biases.pdf}
  \caption{}
  \label{fig:distance-bias}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[page=5, scale=0.25]{biases.pdf}
  \caption{}
  \label{fig:GC-bias}
\end{subfigure}
\begin{subfigure}{\textwidth}
  \centering
  \includegraphics[scale=0.6]{100logo-prob-edited.pdf}
  \caption{}
  \label{fig:logos}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[scale=0.25, page=2]{insertion-position-bias.pdf}
  \caption{}
  \label{fig:insertion-position-bias-ess}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[scale=0.25, page=4]{insertion-position-bias.pdf}
  \caption{}
  \label{fig:insertion-position-bias-ben}
\end{subfigure}
\caption{(a)The distance of the genes from DnaA gene normalised by the lengths of the genomes versus NPEQ. The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance. The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias. (b) G-C content of genes against their NPEQ values. The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias. (c) Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome. The height of each character shows the relative frequency of that character. (d) and (e) The plots show the average insertion index in percentiles of all essential genes (d) and beneficial losses (e). The genes are divided into 3 segments: 5\% of the genes on the $5^\prime$ end, 20\% of the genes on the $3^\prime$ end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
\label{fig:biases}
\end{figure}

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=2, scale=0.22]{biases.pdf}&
%\includegraphics[page=6, scale=0.22]{biases.pdf}&
%\includegraphics[page=10, scale=0.22]{biases.pdf}\\
%\includegraphics[page=14, scale=0.22]{biases.pdf}&
%\includegraphics[page=18, scale=0.22]{biases.pdf}&
%\includegraphics[page=22, scale=0.22]{biases.pdf}\\
%\includegraphics[page=26, scale=0.22]{biases.pdf}&
%\includegraphics[page=30, scale=0.22]{biases.pdf}&
%\includegraphics[page=34, scale=0.22]{biases.pdf}\\
%\includegraphics[page=38, scale=0.22]{biases.pdf}&
%\includegraphics[page=42, scale=0.22]{biases.pdf}&
%\includegraphics[page=46, scale=0.22]{biases.pdf}\\
%\includegraphics[page=50, scale=0.22]{biases.pdf}&
%\includegraphics[page=54, scale=0.22]{biases.pdf}&
%\includegraphics[page=58, scale=0.22]{biases.pdf}\\
%\end{tabular}
%\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes. The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance. The red curves show the loess curve when the smoothness parameter is $0.2$.}
%\label{fig:distance-bias}
%\end{figure*}
\subsubsection{Preferred nucleotide composition bias}
Another concern while inferring essentiality from transposon insertion data is that transposon insertion is biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposon insertion is biased towards, instead of their essentiality level. During Tn5 transposition, a sequence of 9 nucleotides is duplicated. Lodge et al. \cite{lodge_transposon_1988} showed that these duplicated regions have G-C pairs at two ends and replacing these G-C pairs with A-T pairs reduces the number of transposon insertions by more than fivefold. Goryshin et al. \cite{goryshin_tn5/is50_1998} reported the palindromic sequence A-GNTYWRANC-T as the consensus target site for Tn5 transposition where the 9 letters in middle show the consensus sequence in the duplicated region. %They have also showed that insertion sites are usually grouped together and have 5 base pairs distance from each other.
In some other research a similar consensus motif has been found for Tn5 \cite{canals_high-throughput_2012} which is CGCGCA-GTTYWRAAC-TGCGCG. Others \cite{green_insertion_2012,rubin_essential_2015} have not found such a sequence but shown that the duplicated regions are G-C rich. We used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from duplicated regions and 10 nucleotides flanking the 100 top most frequent insertion sites in each genome. The results in Figure \ref{fig:logos} show the consensus motif that we have found is very similar to \cite{green_insertion_2012} and the only difference is in positions 3 and 7 within the duplicated region.

%\begin{figure*}
%\includegraphics[scale=0.9]{100logo-prob-edited.pdf}
%\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome. The height of each character shows the relative frequency of that character.}
%\label{fig:logos}
%\end{figure*}

The other possible source of bias is if transpositions are more inclined to G-C or A-T rich regions. Rubin et al. \cite{rubin_essential_2015} have reported that the number of Tn5 insertions rises with the increase of G-C content and Green et al. \cite{green_insertion_2012} have shown that the highest number of insertions occur in high G-C content regions. On the other hand, Langridge et al. \cite{langridge_simultaneous_2009} have seen an increase in the number of Tn5 insertions in 40\% G-C content. In Figure \ref{fig:GC-bias}, we plotted the G-C content of genes versus their NPEQ values. The red curves are loess curves with smoothness parameter 0.2 and the green line shows no bias. %In most cases (BW25113, EC958, ENC, ETEC, NCTC13441, ROD, SEN, STM, and t)
NPEQ decreases gradually (which means an increase in the number of insertions) as G-C content decreases and then somewhere between 40\% and 50\% G-C content, NPEQ starts to rise again in almost all genomes. %In three cases (SL1344, SL3261, and STMMW) NPEQ loess curve starts to fall off again at about 60\% G-C content which is probably due to a small number of genes with low NPEQ values in that region. The loess curve for BN373 is almost flat. In CS17, the loess curve goes down until it reaches the minimum value (at about 48\%) and then remains unchanged and in ERS227112, the NPEQ value is low at the beginning and then it increases at around 50\% G-C content.
In the region where most of the genes are packed, the loess curve is almost flat. On the left side of this flat region, there are genes with different G-C content which are enriched in mobile genetic elements. So, we expect to see more insertions (smaller NPEQ values) in this region. We have low NPEQ values between 40\% and 50\% G-C content which is expected. However, in most cases when we have less than 40\% G-C content, NPEQ value is high. A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions. This has been shown for Tn10 transposon \cite{kimura_nucleoid_2016}, but not for Tn5 transposon, yet. Overall, the results are consistent with Langridge et al. \cite{langridge_simultaneous_2009}. The G-C content of the most of the genes with large NPEQ values is between 50\% and 60\% which is inconsistent with Green et al. \cite{green_insertion_2012} as this region contains most of the genes and is not considered as a high G-C content region but rather an average G-C content region. %Even though most of the regression lines are consistent with Rubin et al. results \cite{rubin_essential_2015}, the figures show that a linear regression cannot predict G-C bias as its slope changes dramatically in different datasets.

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=1, scale=0.23]{biases.pdf}&
%\includegraphics[page=5, scale=0.23]{biases.pdf}&
%\includegraphics[page=9, scale=0.23]{biases.pdf}\\
%\includegraphics[page=13, scale=0.23]{biases.pdf}&
%\includegraphics[page=17, scale=0.23]{biases.pdf}&
%\includegraphics[page=21, scale=0.23]{biases.pdf}\\
%\includegraphics[page=25, scale=0.23]{biases.pdf}&
%\includegraphics[page=29, scale=0.23]{biases.pdf}&
%\includegraphics[page=33, scale=0.23]{biases.pdf}\\
%\includegraphics[page=37, scale=0.23]{biases.pdf}&
%\includegraphics[page=41, scale=0.23]{biases.pdf}&
%\includegraphics[page=45, scale=0.23]{biases.pdf}\\
%\includegraphics[page=49, scale=0.23]{biases.pdf}&
%\includegraphics[page=53, scale=0.23]{biases.pdf}&
%\includegraphics[page=57, scale=0.23]{biases.pdf}\\
%\end{tabular}
%\caption{The plots show G-C content of genes against their NPEQ values. The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows the linear regression line.}
%\label{fig:GC-bias}
%\end{figure*}

\subsubsection{Positional bias within genes}
Some research has indicated that himar1 transposons are more probable to get inserted into the two ends of a gene compared to the middle \cite{griffin_high-resolution_2011}. We have tested this hypothesis using our TraDIS data. We divided every gene into 100 fragments with equal lengths (percentiles) and calculated the mean insertion index for each percentile. Insertion index is calculated using $\frac{\frac{n_p}{l_p}}{\frac{n_g}{l_g}}$, where $n_p$ is the number of insertion sites in a specific percentile, $l_p$ is the length of that percentile, $n_g$ is the number of insertion sites in the whole genome and $l_g$ is genome length. Mean insertion index for each percentile is calculated by averaging over all insertion indices for that specific percentile of genes. We saw almost no bias towards any location when considering all genes together (Fig.\@ \ref{fig:insertion-position-bias} {\color{red}SUPPLEMENTARY}). We studied the bias in three different groups of genes: essential genes which have no or just a few insertions, non-essential genes which have an intermediate number of insertions, and beneficial losses which have a high number of insertions. The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the $5^\prime$ and $3^\prime$ ends (Fig.\@ \ref{fig:insertion-position-bias-ess}) while it is the opposite in beneficial losses (Fig.\@ \ref{fig:insertion-position-bias-ben}). High number of insertions at the $3^\prime$ end of essential genes implies that the functional domains are located before the insertions and the insertions are not interfering with them. On the other hand, high number of insertions at the $5^\prime$ end of the essential genes indicates there might be alternative start codons in the $5^\prime$ end or it might be because of annotation errors that have predicted the start codon in an incorrect place before the actual start codon.

>>>>

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

@@

% this is required to include graphics
\usepackage{graphicx}
\graphicspath{{../figures/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar

% document begins here
\begin{document}
In this document, I have compared different essential gene predictors. These include BioTraDIS \cite{barquist_tradis_2016} (blue), log fold changes from Monte Carlo method \cite{turner_essential_2015} (green), P-values from running DESeq and Monte Carlo sampling (red), the ratio between the length of the longest uninterruped segment of the gene and the length of the gene (orange), and the average distance between insertion sites. The last two methods are selected from \cite{freed_combining_2016} Figure S3. To calculate one tailed P-values from two tailed P-values obtained from DESeq, I have divided the P-values by two, and if their corresponding logFC value was less than zero, pval = 1 - pval. The set of essential genes are obtained by comparing the genes in each strain to the essential genes in E.coli K-12 from ecogene study \cite{zhou_ecogene_2013}. The results are shown in Figure~\ref{fig:roc}
\begin{figure*}
\centering
\begin{tabular}{c c}
\includegraphics[page=1, scale=0.5]{essential-call-comparison-NCTC13441.pdf} &
\includegraphics[page=1, scale=0.5]{essential-call-comparison-ETEC.pdf} \\
\includegraphics[page=1, scale=0.5]{essential-call-comparison-CS17.pdf} &\\
\end{tabular}
\caption{The comparison of different methods for calling essential genes in three E.coli strains.}
\label{fig:roc}
\end{figure*}

In Figure~\ref{fig:dotplot}, I have compared the logodds value obtained from BioTraDIS to both logFC and P-values from Monte Carlo method for CS17 strain.

\begin{figure*}
\centering
\begin{tabular}{c c}
\includegraphics[page=2, scale=0.5]{essential-call-comparison-CS17.pdf} &
\includegraphics[page=3, scale=0.5]{essential-call-comparison-CS17.pdf} \\
\end{tabular}
\caption{The figure shows the results of BioTraDIS vs. logFC from Monte Carlo method at left and BioTraDIS vs. DESeq P-values at right.}
\label{fig:dotplot}
\end{figure*}

In Figure~\ref{fig:hist}, I have plotted the distribution of each method. The essential and non-essential genes are well separated in Monte Carlo logFC, largest uninterrupted fraction, and mean distance between inserts. However, it is hard to distinguish between non-essential genes and beneficial losses in all these methods.

\begin{figure*}
\centering
\begin{tabular}{c c}
\includegraphics[page=4, scale=0.5]{essential-call-comparison-CS17.pdf} &
\includegraphics[page=5, scale=0.5]{essential-call-comparison-CS17.pdf} \\
\includegraphics[page=6, scale=0.5]{essential-call-comparison-CS17.pdf} &
\includegraphics[page=7, scale=0.5]{essential-call-comparison-CS17.pdf} \\
\end{tabular}
\caption{The figure shows the distribution of Monte Carlo logFC, Monte Carlo P-value, log2(largest uninterrupted fraction), and log2(mean distance between inserts)}
\label{fig:hist}
\end{figure*}

So far, Monte Carlo with logFC values is the best method. Interestingly, even though the mean distance between inserts and largest uninterrupted fraction methods are very simple, they work pretty well.

\bibliography{references}

%This defines the bibliographies style. Search online for a list of available styles.
\bibliographystyle{apalike}
\end{document}
